Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-05-26 elsevier_MANUSCRIPT.qmd
bulk edits from reviewers

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019]. Despite supporting diverse ecosystems and human populations, dryland regions face mounting hydrologic challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. Groundwater serves as a vital resource in drylands for sustaining ecological functions and supporting human livelihoods [@scanlon2006; @yao2018].

@@

The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow volume to the long-term total streamflow volume expressed as a percentage. BFI serves as a normalized measure of groundwater contribution interannually or between basins. BFI is determined by hydrograph separation and is influenced by the climate and physiographic characteristics of a catchment [@neff2005; @beck2013; @singh2018]. Between catchments, base flow fluctuates according to changes in the moisture content of the vadose zone, influenced by varying levels of evapotranspiration and aquifer storage dynamics [@bosch2017]. Since BFI calculations rely on instrumented stream records, it remains unknown for ungauged catchments, which encompass most of the earth's land surface [@fekete2007]. Addressing this information gap is integral to approaching a comprehensive understanding of groundwater dynamics globally.

Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks. To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods. @ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks (ANN) to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.

@@

Despite advancements, existing large-scale studies have limited applicability to the unique hydrogeologic conditions in arid and semi-arid regions like Arizona, leaving uncertainty regarding groundwater-surface water interactions in these critically water-stressed areas. This study addresses this crucial gap by developing a regionally tailored machine learning model specifically designed to estimate BFI for Arizona's ungauged basins. Utilizing hydrogeologic characteristics and hydroclimatic data, we estimate long-term mean BFI (1991–2020) in ungauged catchments, filling spatial gaps in the sparse streamgage network. Furthermore, regional trends in base flow and BFI at instrumented sites are analyzed alongside coincident climate trends in precipitation, reference evapotranspiration (ETO), and temperature. The outcomes offer novel insights into groundwater contributions to streamflow, supporting water resource management in an area increasingly vulnerable to drought and climate change impacts, and contributing broadly to the understanding of hydrologic processes in dryland environments.

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In summer, localized and intense convective storms stem from the North American Monsoon; in winter, orographic precipitation is delivered by Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. Though winter precipitation provides only 30% of annual average precipitation, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Directly measuring base flow and BFI presents unique challenges [@eckhardt2008]. The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013]. However, many methods have been developed to estimate these values. These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990]. These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study. Comparisons of various base-flow separation techniques have been made in previous studies [e.g. @eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.

@@

@@

Our initial training dataset comprised 7,724 observations (annual observations of BFI at streamgages) across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources. The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek, which have similar highly-karstic, snowmelt-driven recharge areas.

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].

@@

#### Trends in Base Flow and BFI {#sec-bfi-trends}

Trends in BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends. Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test. Statistically significant trends were observed in both metrics, with a 72.2% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.

@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area. Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites. No consistent regional patterns are evident.

@@

Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

@@

Coincident trends are defined as those in which base flow or BFI trends match in direction with climate variable trends over the same period. Results are summarized in @tbl-coincident-trends. This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016]. Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature. Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.

Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).

|           | **Climate Variable** | **Coincidence Percentage (%)** |
|-----------|:--------------------:|:------------------------------:|
| BFI       |        ET~O~         |             44.88              |
|           |    Precipitation     |             53.17              |
|           |     Temperature      |             47.32              |
| Base Flow |        ET~O~         |             55.12              |
|           |    Precipitation     |             64.88              |
|           |     Temperature      |             39.02              |

: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}

@@

Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted . The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values indicate strong model performance across the full dataset (R^2^ = 0.764). The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.

Model performance metrics across various classifications are summarized in @tbl-performance. These metrics show consistent model performance across spatial and climatic classifications. However, the negative pbias values across all classifications indicate a systematic underprediction of BFI. Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those contexts.

@@

: Performance of model predictions for BFI for all sites split by various classifications. n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias. {#tbl-performance}

@@

The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting long-term BFI is basin elevation. While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].

Land cover and land use predictors also play a significant role in BFI estimation. Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence. Similarly, hydrologic soil types show distinct trends in their impact on BFI. Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI. In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.

![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/updated/BFI_HUC8_20250404.png){#fig-bfi-huc}

@@

This study addresses these monitoring gaps through an analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes. Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations. High BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim. Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments. A trend of high BFI in upstream reaches to lower BFI in downstream reaches [@fig-instrumented-bfi] likely indicates a transition from zones of groundwater discharge to zones of groundwater recharge and may indicate a transition from gaining stream reaches to losing stream reaches [@winter2007].

Analysis of trends at instrumented sites revealed a predominance of declining BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins. These negative trends likely reflect intensified climate stress, highlighting areas most vulnerable to groundwater depletion and surface-water scarcity. Precipitation was identified as the primary driver influencing base-flow variability, with temperature and evapotranspiration trends adding complexity. The weaker correlation between temperature and BFI trends suggests that temperature impacts are moderated by other environmental factors, underscoring the need for integrated climate and land-management strategies to protect groundwater-dependent ecosystems.

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

@@

The long-term average BFI across Arizona is approximately 0.32, highlighting that groundwater discharge significantly contributes to surface water flows but exhibits substantial spatial variability. Our machine learning model (XGBoost) effectively predicts BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics. Regions such as the Grand Canyon and Mogollon Rim exhibit the highest BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.

Analysis of temporal trends indicates a prevailing pattern of declining base flow and BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas. Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity. The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios and other stresses to aquifers, such as increased groundwater pumping.

>>>>

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, cover 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019]. Despite supporting diverse ecosystems and human populations, dryland regions face mounting hydrologic challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. Groundwater serves as a vital resource in drylands for sustaining ecological functions and supporting human livelihoods [@scanlon2006; @yao2018].

@@

The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow volume to the long-term total streamflow volume expressed as a percentage. BFI serves as a normalized measure of groundwater contribution interannually or between basins. BFI is determined by hydrograph separation and is influenced by the climate and physiographic characteristics of a catchment [@neff2005; @beck2013; @singh2018]. Between catchments, base flow fluctuates according to changes in the moisture content of the vadose zone, influenced by varying levels of evapotranspiration and aquifer storage dynamics [@bosch2017]. Since BFI calculations rely on instrumented stream records, it remains unknown for ungauged catchments, which encompass most of the earth's land surface [@fekete2007]. Addressing this information gap is integral to approaching a comprehensive understanding of groundwater dynamics globally. This study addresses this gap by analyzing long-term BFI patterns across Arizona using instrumented streamflow records and applying a machine learning model to predict BFI in ungauged catchments.

Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks. To tackle the challenge of quantifying base-flow indices in ungauged catchments, numerous studies have applied both regression and machine learning methods. @ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks (ANN) to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.

@@

Despite advancements, existing large-scale studies have limited applicability to the unique hydrogeologic conditions in arid and semi-arid regions like Arizona, leaving uncertainty regarding groundwater-surface water interactions in these critically water-stressed areas. This study addresses this crucial gap by developing a regionally tailored machine learning model specifically designed to estimate BFI for Arizona's ungauged basins. Utilizing hydrogeologic characteristics and hydroclimatic data, we estimate long-term mean BFI (1991–2020) in ungauged catchments, filling spatial gaps in the sparse streamgage network. Furthermore, regional trends in base flow and BFI at instrumented sites are analyzed alongside coincident climate trends in precipitation, reference evapotranspiration (ETO), and temperature. The outcomes offer novel insights into groundwater contributions to streamflow, supporting water resource management in an area increasingly vulnerable to drought and climate change impacts, and contributing broadly to the understanding of hydrologic processes in dryland environments. The objectives of this study are to (1) quantify long-term spatial and temporal trends in BFI across Arizona, and (2) develop a predictive model using eXtreme Gradient Boosting (XGBoost) to estimate BFI in ungauged basins based on hydroclimatic and geospatial variables.

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In summer, localized and intense convective storms stem from the North American Monsoon; in winter, orographic precipitation is delivered by Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. Though winter precipitation accounts for only 30% of annual precipitation totals, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Directly measuring base flow and base-flow index (BFI) presents unique challenges [@eckhardt2008]. Base flow is a physical hydrologic process representing groundwater contributions to streamflow, while BFI is a dimensionless, statistical metric that estimates the proportion of streamflow derived from base flow. Because BFI must be derived using a base-flow separation technique, its value is inherently method-dependent and sensitive to the separation approach used [@beck2013]. A variety of techniques have been developed to estimate base flow, including tracer studies [@gonzales2009], graphical interpolation methods [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990]. Each of these methods varies in its applicability depending on spatial scale, record length, and study objectives. Although the choice of separation method and parameterization introduces some subjectivity, these filters have been shown to yield reliable estimates when applied consistently within a study domain [@chapman_1999; @eckhardt2005; @instituteofhydrology1980; @ayers2022]. Numerous studies have compared these separation techniques [e.g., @eckhardt2005; @eckhardt2008; @nathan1990]; however, this study does not evaluate the relative performance of different methods.

@@

Two BFI metrics were used in this study: (1) annual BFI, calculated for each calendar year using daily streamflow records, and (2) long-term BFI, defined as the average of annual BFI values over the period of record at each site. Annual BFI reflects interannual variability in base-flow contributions, while long-term BFI represents a time-averaged metric of groundwater influence. The machine learning model was trained to predict annual BFI values, which were then aggregated to produce long-term BFI estimates.

@@

Our training dataset consisted of 7,724 site-year observations of annual BFI, where each observation represents a single water year at a given streamgage. These observations were paired with 45 predictor variables derived from basin characteristics and climate data. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources. The highest long-term BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. High long-term BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek, which have similar highly-karstic, snowmelt-driven recharge areas.

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate low long-term BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].

@@

#### Trends in BFI {#sec-bfi-trends}

Trends in long-term BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends. Long-term BFI trends were assessed at all instrumented sites using the Mann–Kendall test. As BFI quantifies the proportion of streamflow attributed to base flow, its long-term trends inherently reflect changes in subsurface contributions relative to total flow. A 72.2% coincidence rate between significant base flow and BFI trends supports the interpretation that changes in BFI are largely driven by underlying base-flow dynamics. Statistically significant long-term BFI trends were observed across numerous sites, with spatial variation highlighting regional differences in hydrologic response.

@fig-instrumented-trend illustrates the spatial variation in long-term BFI trends across the study area. Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites. No consistent regional patterns are evident.

@@

Statistically significant decreasing trends in long-term BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining long-term BFI. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to BFI declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

@@

Coincident trends are defined as those in which long-term BFI trends match in direction with climate variable trends over the same period. Results are summarized in @tbl-coincident-trends. This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016]. Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature. Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.

Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends of long-term BFI in relation to coincident trends in climate variables. Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in BFI (@tbl-coincident-trends).

|     | **Climate Variable** | **Coincidence Percentage (%)** |
|-----|:--------------------:|:------------------------------:|
| BFI |        ET~O~         |             44.88              |
|     |    Precipitation     |             53.17              |
|     |     Temperature      |             47.32              |

: Coincident trends of climate variables (ET~O~, precipitation, temperature) with BFI trends. {#tbl-coincident-trends}

@@

Predicted values of annual BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted . The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values indicate strong model performance across the full dataset (R^2^ = 0.764). The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.

Model performance metrics across various classifications are summarized in @tbl-performance. These metrics show consistent model performance across spatial and climatic classifications. However, the negative pbias values across all classifications indicate a systematic underprediction of annual BFI. Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those contexts.

@@

: Performance of model predictions for annual BFI for all sites split by various classifications. n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias. {#tbl-performance}

@@

The predictors used to estimate annual BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values. The most influential feature for predicting annual BFI is basin elevation. While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013]. The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration. Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].

Land cover and land use predictors also play a significant role in annual BFI estimation. Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence. Similarly, hydrologic soil types show distinct trends in their impact on BFI. Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI. In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc. Basins with high long-term BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher long-term BFI values. In contrast, low long-term BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.

![Predicted long-term BFI values for 8-digit HUC (1991-2020). Bolded HUCs indicate those without a streamgage.](images/BFI_HUC8_20250525.png){#fig-bfi-huc}

@@

This study addresses these monitoring gaps through an analysis of observed BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes. Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations. High long-term BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim. Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments. A trend of high BFI in upstream reaches to lower BFI in downstream reaches [@fig-instrumented-bfi] likely indicates a transition from zones of groundwater discharge to zones of groundwater recharge and may indicate a transition from gaining stream reaches to losing stream reaches [@winter2007].

Analysis of trends at instrumented sites revealed a predominance of declining long-term BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins. These negative trends likely reflect intensified climate stress, highlighting areas most vulnerable to groundwater depletion and surface-water scarcity. Precipitation was identified as the primary driver influencing base-flow variability, with temperature and evapotranspiration trends adding complexity. The weaker correlation between temperature and BFI trends suggests that temperature impacts are moderated by other environmental factors, underscoring the need for integrated climate and land-management strategies to protect groundwater-dependent ecosystems.

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate annual BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

@@

The long-term average BFI across Arizona is approximately 0.32, highlighting that groundwater discharge significantly contributes to surface water flows but exhibits substantial spatial variability. Our machine learning model (XGBoost) effectively predicts annual BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics. Regions such as the Grand Canyon and Mogollon Rim exhibit the highest long-term BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.

Analysis of temporal trends indicates a prevailing pattern of declining long-term BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas. Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity. The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios and other stresses to aquifers, such as increased groundwater pumping.
